---
title: "Project 1 Data Analysis Prototype"
author: "Jonathon Roach"
output: word_document
---

## Business Understanding

The Belk Foundation's website says, "Our goal is to empower today’s workforce by creating pathways to and through postsecondary education for underrepresented students". For the sake of this analysis, we assume that better performing schools have better outcomes in postsecondary education. With North Carolina's rapidly-changing demographics, it is important to take into consideration schools' unique needs when allocating funds to strategic investment initiatives. Here, we explore where funding can be best applied based to educational achievement data.

In order to achieve this goal, we will explore through visual and mathematical modeling which features best predict the School Performance Grade (SPG), a measure of a school's overall success based on test scores and growth measures. 

# Visualize Attributes

Firstly, let's take a look at the data we're working with. 

```{r loading libraries}
library(dplyr)
library(stringr)
library(ggplot2)
set.seed(123)

```


```{r data import, cache=TRUE, warning=FALSE}

schools <- read.csv('PublicSchools2014to2017_YZ.csv', stringsAsFactors = F)

```

It is important to understand the types of schools that are relevant to this analysis. Let's first take a look at school type and see if there are any anomolies (schools for children with disabilities, for example).  

```{r look at school types}

table(schools$school_type_txt)

```

"Regular" schools seem to make up the majority of the dataset, so we will focus on those for now. Perhaps we can come back to the the other school types later and analyze them separately. 

```{r filtering to regular schools}

schools <- schools %>% filter(school_type_txt == "Regular School")

```

Let's now take a look at schools by the age ranges of their students: elementary, middle, high, and various combinations of the three. 

```{r recoding category_cd}

#Re-coding the levels of the variable 'category_cd' so they're easier to read
#in reports. 


schools$category_cd <- 
case_when(
  
  schools$category_cd == 'A' ~ 'Elem./Mid./High Together',
  schools$category_cd == 'E' ~ 'Elementary School',
  schools$category_cd == 'H' ~ 'High School',
  schools$category_cd == 'I' ~ 'Elem./Mid. Together',
  schools$category_cd == 'M' ~ 'Middle School',
  TRUE ~ 'Mid./High Together',
  
)


table(schools$category_cd)

```

Some of these categories have very little representation in the data. For now, we'll remedy this by lumping the combo groups together. 

```{r lumping combo categories}

schools$category_cd <-
  ifelse(str_detect(schools$category_cd, "/"),
         "Combo",
         schools$category_cd)

table(schools$category_cd)

```

We now need to somehow get at the idea of demographic composition of schools. Dr. Drew and his capstone groups have shown that classifying schools as majority-minority when they are composed of >50% non-white students highlights meaningful differences in school performance (likely due the fact that demograhpics can serve as a stand-in for economic measures). Let's take the same approach.

```{r defining majority-minority}

schools$MinorityOverallPct <- schools$MinorityMalePct + schools$MinorityFemalePct
schools$Majority_Minority <- ifelse(schools$MinorityOverallPct > .5, 1, 0)

```

Now let's do a univariate analysis on the impact of attendance rates on school performance grade. The expectation prior to doing the analysis is that poor attendance rates result in poor school performance.

```{r plotting attendance vs. score}

schools %>%
  ggplot(aes(avg_daily_attend_pct, SPG.Score)) +
    geom_point() +
    facet_wrap(~Year) + 
    geom_smooth(method = "lm", se=F) +
    labs(x="Average Daily Attendance %", y = "SPG Score (0-100)") +
    ggtitle("SPG Score vs. Average Daily Attendance %")

```


Beyond confirming our suspicions, this graph also reveals an interesting quirk about the data. For each year, there is a group of schools that received a zero for their SPG score despite having high average attendance. Manually inspecting these records reveals that these are all elementary schools. We will keep this in mind as we conduct further analyses.

Some people theorize that smaller class sizes are associated with better schools. Let's test this.

```{r plotting school size vs. score}

schools %>%
  filter(category_cd == "High School") %>%
  ggplot(aes(student_num, SPG.Score)) +
    geom_point() +
    facet_wrap(~Year) +
    labs(x="Number of Students at School", y="SPG Score") +
    ggtitle("SPG Score by School Size", subtitle = "High School Students Only")
  
schools %>%
  filter(category_cd == "Middle School") %>%
  ggplot(aes(student_num, SPG.Score)) +
    geom_point() +
    facet_wrap(~Year) +
    labs(x="Number of Students at School", y="SPG Score") +
    ggtitle("SPG Score by School Size", subtitle = "Middle School Students Only")


schools %>%
  filter(category_cd == "Elementary School") %>%
  ggplot(aes(student_num, SPG.Score)) +
    geom_point() +
    facet_wrap(~Year) +
    labs(x="Number of Students at School", y="SPG Score") +
    ggtitle("SPG Score by School Size", subtitle = "Elementary School Students Only")

```

While this isn't a good measure of class size (we have no way of measuring student/teacher ratio. Just raw number of students), it's interesting to observe that larger public schools generally receive higher grades than smaller ones. It could be the case that here the size of the school is a stand-in for the rurality of the school. In rural areas, schools will be smaller, and at the same time, there will be fewer opportuntities for students. What happens when we color by the majority-minority variable? (looking for trends in large inner-city schools)

```{r plotting school size vs. score again}

schools %>%
  filter(category_cd == "High School") %>%
  ggplot(aes(student_num, SPG.Score, group = factor(Majority_Minority), color=factor(Majority_Minority))) +
    geom_smooth(se=F) +
    facet_wrap(~Year) +
    labs(x="Number of Students at School", y="SPG Score") +
    ggtitle("SPG Score by School Size", subtitle = "High School Students Only")
  
schools %>%
  filter(category_cd == "Middle School") %>%
  ggplot(aes(student_num, SPG.Score, group = factor(Majority_Minority), color=factor(Majority_Minority))) +
    geom_smooth(se=F) +
    facet_wrap(~Year) +
    labs(x="Number of Students at School", y="SPG Score") +
    ggtitle("SPG Score by School Size", subtitle = "Middle School Students Only")


schools %>%
  filter(category_cd == "Elementary School") %>%
  ggplot(aes(student_num, SPG.Score, group = factor(Majority_Minority), color=factor(Majority_Minority))) +
    geom_smooth(se=F) +
    facet_wrap(~Year) +
    labs(x="Number of Students at School", y="SPG Score") +
    ggtitle("SPG Score by School Size", subtitle = "Elementary School Students Only")

```

This is perhaps consistent with our hypothesis stated earlier, but not necessarily. High schools exhibit a particularly interesting trend. There seems to be some benefit to being either an exceptionally small or large high school, but being medium sized seems to be detrimental to the overall school quality. Not sure why this would be.

Now let's turn our attention to teacher hiring. It would be reasonable to hypothesize that having teachers with more experience would result in better educational quality. Let's test it out. 

```{r investigating teacher hiring}

schools %>%
  filter(category_cd == "High School") %>%
  ggplot(aes(tchyrs_0thru3_pct, SPG.Score)) +
    geom_point() +
    geom_smooth(method = "lm", se=F) +
    facet_grid(Year~Majority_Minority) +
    labs(x="Proportion of Teachers With 0-3 Years Experience", y="SPG Score") +
    ggtitle("SPG Score by Teacher Experience", subtitle = "High School Students Only")

schools %>%
  filter(category_cd == "Middle School") %>%
  ggplot(aes(tchyrs_0thru3_pct, SPG.Score)) +
    geom_point() +
    geom_smooth(method = "lm", se=F) +
    facet_grid(Year~Majority_Minority) +
    labs(x="Proportion of Teachers With 0-3 Years Experience", y="SPG Score") +
    ggtitle("SPG Score by Teacher Experience", subtitle = "Middle School Students Only")

schools %>%
  filter(category_cd == "Elementary School") %>%
  ggplot(aes(tchyrs_0thru3_pct, SPG.Score)) +
    geom_point() +
    geom_smooth(method = "lm", se=F) +
    facet_grid(Year~Majority_Minority) +
    labs(x="Proportion of Teachers With 0-3 Years Experience", y="SPG Score") +
    ggtitle("SPG Score by Teacher Experience", subtitle = "Elementary School Students Only")

```

Does disciplinary policy effect school quality? Let's take a look at suspensions. 

```{r investigation of suspensions}

schools %>%
  filter(category_cd == "High School") %>%
  ggplot(aes(short_susp_per_c_num, SPG.Score)) +
    geom_point() +
    geom_smooth(method = "lm", se=F) +
    facet_grid(Year~Majority_Minority) +
    labs(x="Average Short-Term Suspensions Per 100 Students", y="SPG Score") +
    ggtitle("SPG Score by Short-Term Suspension Rate", subtitle = "High School Students Only")

schools %>%
  filter(category_cd == "Middle School") %>%
  ggplot(aes(short_susp_per_c_num, SPG.Score)) +
    geom_point() +
    geom_smooth(method = "lm", se=F) +
    facet_grid(Year~Majority_Minority) +
    labs(x="Average Short-Term Suspensions Per 100 Students", y="SPG Score") +
    ggtitle("SPG Score by Short-Term Suspension Rate", subtitle = "Middle School Students Only")

schools %>%
  filter(category_cd == "Elementary School") %>%
  ggplot(aes(short_susp_per_c_num, SPG.Score)) +
    geom_point() +
    geom_smooth(method = "lm", se=F) +
    facet_grid(Year~Majority_Minority) +
    labs(x="Average Short-Term Suspensions Per 100 Students", y="SPG Score") +
    ggtitle("SPG Score by Short-Term Suspension Rate", subtitle = "Elementary School Students Only")

```

Short-term suspensions appear to be ineffective means of raising educational quality. 

With the trends we've identified above, let's build a linear regression model to see how feasible it is to predict SPG Score. First we'll start by training the model on data points that make sense using a 60/40 train-test split. Then we'll run cross-validation to see how we did.

```{r}

#recall from earlier we have quite a few points with an SPG score of zero.
#Let's get rid of those.

schools <-  schools %>% filter(SPG.Score != 0)

inds <- sample(nrow(schools), floor(0.6*nrow(schools)), replace = F)
train <- schools[inds,]
test <- schools[-inds,]

mod <- lm(formula = SPG.Score ~ category_cd + Majority_Minority + avg_daily_attend_pct + student_num + tchyrs_0thru3_pct + short_susp_per_c_num, data = train)

#Calculating RMSE
sqrt(sum((test$SPG.Score - predict(mod, newdata = test))^2) / nrow(test))

```

Looks like our model works pretty well. On average, it misses the grade by around 9 points or so. If you run the summary() function on the model, you'll find that all the terms are significant. Nice!

Preliminary conclusions:

The Belk Foundation should invest money into initiatives that involve the following:

1.) Getting students to attend class more often. 
2.) Hiring teachers with > 3 years of experience to teach at underserved schools.
3.) Finding alternative disciniplinary measures to short-term suspensions.
4.) Providing growth opportunities for students that attend schools that are too large for personalized attention, but too small for extensive extracurriculars.
